A Speculative Approach to Parallelization in Particle Swarm Optimization

### 2011-02-14 masters_thesis/si_paper/paper.tex
Reworked a couple of sections in the paper after feedback from Andrew

There have been several parallel implementations of PSO presented in the
literature.  These implementations vary along two axes---what computation is
performed and how it is done.  How the computation is performed involves both
architectural decisions, such as a master-slave versus a fully distributed
architecture, and decisions about synchronicity.  What computation is performed
involves decisions about how to scale the PSO algorithm for parallel
implementations, such as the ratio of particles to processors or the topology
to use.

\subsection{Synchronous vs. Asynchronous PSO}

@@

The contribution of this work deals not with the implementation details of
architecture and synchronicity, but with what should be done with the PSO
equations when many hundreds or thousands of processors are available.  The
main issues that have been addressed in this space are how many particles to
use for a particular number of processors and what communication topology
should be employed.

The number of particles per processor has typically been decided by how long it
takes to evaluate the function being optimized.  When the function takes longer
than a few seconds to evaluate, previous techniques have assigned the number of
particles in the swarm to be the number of processors available~%
\citep{jin-2005-pso-antenna-designs,mcnabb-2009-large-particle-swarms},
advocating using as many processors as possible to get the best performance.
When the function takes less time to evaluate than the TCP/IP stack used to
send interprocessor communication, parallel implementations assign several or
many particles to a single processor~\citep{chu-2006-intelligent-parallel-pso,
chang-2005-parallel-pso-with-communication-strategies}.  Often the processor
only sends information about the best particle it evaluated to other
processors~\citep{mcnabb-2009-large-particle-swarms}.

@@

With regard to increasing the swarm size in PSO, some recent work has suggested
that increasing the swarm size through the course of the optimization process
provides better results than having a set swarm
size~\citep{hsieh-2009-efficient-population-utilization-for-pso,
montes-de-oca-2010-incremental-social-learning-pso}.  However, these results
are based on total number of function evaluations, which, when running in
parallel on expensive functions, is less important than total number of
iterations.  Other work focusing on parallelization has shown that when extra
processors are available they should be used, as performance increases with
swarm size when measuring in terms of number of
iterations~\citep{mcnabb-2009-large-particle-swarms,
jin-2005-pso-antenna-designs}.  If the swarm size were varied throughout the
course of the optimization process, some processors would be sitting idle at
most iterations.

@@

\subsection{Refactoring the PSO Equations}

@@

\subsection{Discussion}

To see the value of this refactoring, suppose that $1000$ processors are
available, and that the evaluation of the objective function takes one hour.
If we only want a swarm of $100$ particles, $900$ of the processors would be
sitting idle for an hour at every iteration, and it would take two hours to run
two iterations.  If instead we perform speculative evaluation, sending each of
the $7$ evaluations in $\{\ofunc(\specpos_{t+1}^{c}) : \casegen \in \caseset\}$
to be computed at the same time as $\ofunc(\ppos_{t})$, we would use $800$ of
the $1000$ processors and perform two iterations in one hour.

>>>>

There have been several parallel implementations of PSO presented in the
literature.  The improvements described in the literature come in two major
areas: innovations in implementation details and innovations in the use of
topology and swarm size to scale PSO to many processors.

\subsection{Innovative Implementations}

There are many ways to parallelize the basic PSO algorithm.  The most
fundamental decision to make in parallel PSO is which parallel architecture to
use.  Several architectures have been proposed, including Master-Slave, fully
distributed (sometimes called ``diffusion''), and reformulating PSO into
Google's MapReduce framework~\citep{belal-2004-parallel-models-for-pso,
mcnabb-2007-parallel-pso-using-mapreduce}.  A somewhat orthogonal
implementation decision when parallelizing PSO is whether to have synchronous
communication or asynchronous communication.

@@

The other area of research in parallelizing PSO deals not with the
implementation details of architecture and synchronicity, but with what should
be done with the PSO equations when many hundreds or thousands of processors
are available.  The main issues that have been addressed in this space are how
many particles to use for a particular number of processors and what
communication topology should be employed.

The number of particles per processor has typically been decided by how long it
takes to evaluate the function being optimized.  When the function takes longer
than a few seconds to evaluate, previous techniques have assigned the number of
particles in the swarm to be the number of processors available~%
\citep{jin-2005-pso-antenna-designs,mcnabb-2009-large-particle-swarms},
advocating using as many processors as possible to get the best performance.
When the function takes less time to evaluate than the TCP/IP stack used to
send interprocessor communication, parallel implementations assign several or
many particles to a single processor~\citep{chu-2006-intelligent-parallel-pso,
chang-2005-parallel-pso-with-communication-strategies}.  Often the processor
only sends information about the best particle it evaluated to other
processors~\citep{belal-2004-parallel-models-for-pso}.

@@

With regard to increasing the swarm size in PSO, some recent work has suggested
that increasing the swarm size throughout the course of the optimization
process provides better results than having a set swarm
size~\citep{hsieh-2009-efficient-population-utilization-for-pso,
montes-de-oca-2010-incremental-social-learning-pso}.  However, these results
focused on serial computation and are based on total number of function
evaluations, which, when running in parallel on expensive functions, is less
important than total number of iterations.  Other work focusing on
parallelization has shown that when extra processors are available they should
be used, as performance increases with swarm size when measuring in terms of
number of iterations~\citep{mcnabb-2009-large-particle-swarms,
jin-2005-pso-antenna-designs}.  If the swarm size were varied throughout the
course of the optimization process, some processors would be sitting idle at
most iterations.

The contribution of our work is in this area of what should be done with the
PSO equations to better utilize a thousand processors when they are available.
In our work we use a synchronous, MapReduce implementation of parallel PSO.
While we use a specific implementation, we describe how speculative evaluation
can be performed in any of the synchronous architectures mentioned in the
previous section.  The adaptation of our methods to asynchronous PSO
parallelization methods should be straightforward, though it is left to future
work.

@@

Our speculative methods require refactoring the PSO equations such that all
possible positions for each particle at iteration $t+1$ can be evaluated in
parallel along with the position of each particle at iteration $t$.  With some
careful bookkeeping, we can then piece together the results of iteration $t+1$
for each particle, thus using extra processors to evaluate two iterations of
the algorithm in the time it takes to evaluate the function once.  As we show
in Sections~\ref{sec:proof} and~\ref{sec:topology}, a wise choice of topology
limits the necessary speculative evaluations to seven per particle.

To see the value of this refactoring, suppose that $1000$ processors are
available, and that the evaluation of the objective function takes one hour.
If we only want a swarm of $100$ particles, $900$ of the processors would be
sitting idle for an hour at every iteration, and it would take two hours to run
two iterations.  If instead we perform speculative evaluation, sending each of
the $7$ possible speculative positions of a particle to be computed at the same
time as its current position, we would use $800$ of the $1000$ processors and
perform two iterations in one hour.

\sec{proof} shows in detail how the PSO equations can be refactored to allow
for speculative evaluation, proving that our methods are correct and that they
exactly reproduce the behavior of the PSO algorithm.  The section also
introduces some notation used later in the paper.  \sec{topology} gives a brief
discussion of how the topology used affects the amount of speculative
computation needed.

\subsection{Refactoring the PSO Equations}
\label{sec:proof}

@@

\subsection{Topology in Speculative Evaluation}
\label{sec:topology}
